---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%", 
  dpi=300
)
```

# cst: climate scenarios toolkit

<!-- badges: start -->
[![Build Status](https://travis-ci.com/earthlab/cst.svg?branch=master)](https://travis-ci.com/earthlab/cst)
[![codecov](https://codecov.io/gh/earthlab/cst/branch/master/graph/badge.svg)](https://codecov.io/gh/earthlab/cst)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
<!-- badges: end -->

The goal of cst is to provide easy climate data access 
([MACA v2](http://www.climatologylab.org/maca.html)) to support 
climate scenario planning.
This package allows you to: 

1. Quickly acquire climate data subsets for a spatial region of interest, with
first class support for US National Parks
2. Summarize climate data at daily timesteps, and compute derived quantities
3. Contrast reference and target time periods to understand differences in 
climate over time, and
4. Easily work with climate data, without having to worry about the details of 
how it is stored or formatted

## Installation

Install the development version of cst from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("earthlab/cst")
```

This package also has some python dependencies. 
We recommend using conda (preferably [miniconda](https://docs.conda.io/en/latest/miniconda.html)) to simplify the installation process. 
Python dependencies can be installed as follows: 

``` r
cst::install_py_deps()
```

## Quickstart guide

To acquire daily precipitation data for Acadia National Park for a subset of 
climate models, you can use the `cstdata()` function:

```{r}
library(ggplot2)
library(cst)

d <- cstdata(park = "Acadia National Park", parameters = "pr", 
             years = c(2020, 2021), models = "CCSM4", scenarios = "rcp85")
```

This gives you a data frame with paths to local climate data files: 

```{r print-d}
str(d)
```

And, you can also summarize the daily data by computing a spatial average over the region of interest: 

```{r summarize-data}
df <- cst_df(d, cores = 2)
str(df)
```

Because this is a data.frame, you can use all of the normal data visualization and processing functionality in R, e.g.,

```{r ggplot-precip, fig.width = 5, fig.height = 2.5}
df %>%
  ggplot(aes(date, pr)) + 
  geom_point() + 
  geom_line(alpha = .1) + 
  xlab("Date") + 
  ylab("Precipitation (mm)") + 
  ggtitle("Acadia National Park, CCSM4, RCP 8.5")
```

### Dive deeper

This is just a small glimpse at what you can do with the cst package.
For more, see [Getting started with the Climate Scenarios Toolkit](https://www.earthdatascience.org/cst/articles/cst-intro.html)

## Meta

* Please [report any issues or bugs](https://github.com/earthlab/cst/issues),
after reading our contribution [guidelines](CONTRIBUTING.md), and the 
[Contributor Code of Conduct](CODE_OF_CONDUCT.md). 
* License: GPL-3
* See `citation("cst")` in R to cite this package in publications. 