---
title: "Climate Futures Toolbox"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2

---
# Welcome to the Climate Futures Toolbox

  This is a package developed as a collaboration between Earth lab and the North Central Climate Adaptation Science Center to help users gain insights from available climate data. This package
includes tools and instructions for downloading climate data via a USGS API and then organizing those
data for visualization and analysis that drive insight.

   This package is currently growing to include better functionality for spatial analyses and more user-friendly features. Thank you for all the wonderful beta tester groups that helped us get the software this far. Please be patient as we update some of the functions and vignette to accommodate more functionality. 

# What you'll find here
This vignette provides a walk-through of a common use case of the cft package, which is, to
help users download, organize, and visualize past and future climate data.
- 1) How to download and install the cft package
- 2) How to see the menu of available data and choose items from that menu
- 3) How to request data from the API using those menu choices
- 4) How to aggregate those data in differeent ways to drive insight.  

## Why write the cft package?

The amount of data generated by downscaled GCMs can be quite large
(e.g., daily data at a few km spatial resolution).
The Climate Futures Toolbox was developed to help users access and use 
smaller subsets. 

Data is acquired from the [Northwest Knowledge Server of the University of
Idaho](http://thredds.northwestknowledge.net:8080/thredds/reacch_climate_CMIP5_macav2_catalog2.html). 

### What you'll need

To get the most out of this vignette, we assume you have: 

- At least 500 MB of disk space
- Some familiarity with ggplot2
- Some familiarity with dplyr (e.g., `filter()`, `group_by()`, and `summarize()`)

## About the data

Global Circulation Models (GCMs) provide estimates of historical and future 
climate conditions. 
The complexity of the climate system has lead to a large number GCMs and it is
common practice to examine outputs from many different models, treating each as 
one plausible future.

Most GCMs are spatially coarse (often 1 degree), but downscaling provides finer
scale estimates. The cft package uses one downscaled climate model called MACA 
(Multivariate Adaptive Climate Analog) Version 2 
([details here](http://www.climatologylab.org/maca.html)).


# Loading the cft package from github
```{r install cft, warning=FALSE, message=FALSE, eval=FALSE}
library(devtools)
install_github("earthlab/cft", force = TRUE)
```

## Attach cft and check the list of available functions
```{r}
#install.packages("sp")
#install.packages("multidplyr")
library(multidplyr)
library(cft)
ls(pos="package:cft")
```

## Look at the documentation for those functions
```{r manual}
?available_data
```



# Use read-only mode to find available data without initiating a full download.
```{r available data, cache=TRUE}
start_time <- Sys.time()
inputs <- cft::available_data()
end_time <- Sys.time()

end_time - start_time
```

# Filter the results from available_data() to specify which data to actually download.
  There are too many data available to download in a single download request. You will need limit your requests to 500mb. This is enough to download a single variable for a single spatial point for the full available time period, but not more than that. If you want to pull multiple variables and/or multiple spatial locations, you will need to submit multiple request for data and merge those tables together after download. We provide examples for both.


## We can look at just the unique variable types to get an idea of what's available 
```{r}
levels(as.factor(inputs$variable_names$Variable))
levels(as.factor(inputs$variable_names$`Variable abbreviation`))

levels(as.factor(inputs$variable_names$Scenario))
levels(as.factor(inputs$variable_names$`Scenario abbreviation`))

levels(as.factor(inputs$variable_names$Scenario))
levels(as.factor(inputs$variable_names$`Scenario abbreviation`))

levels(as.factor(inputs$variable_names$Model))
levels(as.factor(inputs$variable_names$`Model abbreviation`))
```

## But we prefer to use the table version so we can easily select combinations of variables that we want to pull together.
  Here you will use tidy notation to filter the avalable_data table to only include the entries that you would like to download. 

## Filter variable names
This filter includes all of the climate models, all of the scenarios, and 5 variables. It is a big request.
```{r filter variables many, cache=TRUE}
input_variables <- inputs$variable_names %>% 
  filter(Variable %in% c("Maximum Relative Humidity", 
                       "Maximum Temperature", 
                       "Minimum Relative Humidity",          
                       "Minimum Temperature",                 
                       "Precipitation")) %>% 
  filter(Scenario %in% c( "RCP 4.5", "RCP 8.5")) %>% 
  filter(Model %in% c(
    "Beijing Climate Center - Climate System Model 1.1",
    "Beijing Normal University - Earth System Model",
    "Canadian Earth System Model 2",                                                                
  "Centre National de Recherches Météorologiques - Climate Model 5",                              
  "Commonwealth Scientific and Industrial Research Organisation - Mk3.6.0",                       
  "Community Climate System Model 4",                                                             
  "Geophysical Fluid Dynamics Laboratory - Earth System Model 2 Generalized Ocean Layer Dynamics",
  "Geophysical Fluid Dynamics Laboratory - Earth System Model 2 Modular Ocean",                   
  "Hadley Global Environment Model 2 - Climate Chemistry 365 (day) ",                             
 "Hadley Global Environment Model 2 - Earth System 365 (day)",                                   
 "Institut Pierre Simon Laplace (IPSL) - Climate Model 5A - Low Resolution",                     
 "Institut Pierre Simon Laplace (IPSL) - Climate Model 5A - Medium Resolution",                  
 "Institut Pierre Simon Laplace (IPSL) - Climate Model 5B - Low Resolution",                     
 "Institute of Numerical Mathematics Climate Model 4",                                           
 "Meteorological Research Institute - Coupled Global Climate Model 3",                           
 "Model for Interdisciplinary Research On Climate - Earth System Model",                         
 "Model for Interdisciplinary Research On Climate - Earth System Model - Chemistry",             
 "Model for Interdisciplinary Research On Climate 5",                                            
 "Norwegian Earth System Model 1 - Medium Resolution"  )) %>%
  
  pull("Available variable")

input_variables
```


# Establish area of interst (AOI) by bounding box
You will need to specify an appropriate area of interest for downloading spatial data. We utilize the functionality of open street map to make it fast and easy to download shapefiles for any area of interest. Here I am using the open street map (OSM) protocol to retrive a shapefile for Hot Springs National Park. I choice this park because it is the smallest national park. 

This chunk of code is first using the getbb() function to retrive a bounding box matching a plain language search request. That bounding box is then fed into the API call that retrives the data and converts it into an r-specific spatial data format called sf.

OSM uses a 'key' and 'value' system for organizing requests. You may want to spend some time familirizing yourself with the immense library of data you can access with this system. https://wiki.openstreetmap.org/wiki/Tags
```{r bounding box small, cache=TRUE}
bb <- getbb("Hot Springs")
my_boundary <- opq(bb) %>% 
  add_osm_feature(key = "boundary", value = "national_park") %>% 
osmdata_sf() 

my_boundary
```
```{r bounding box colorado, cache=TRUE}
bb <- getbb("Colorado")
my_boundary <- opq(bb) %>% 
  add_osm_feature(key = "admin_level", value = "4") %>% 
osmdata_sf() 

my_boundary
```


Here you re-calibrate the bounding box relative to the actual shapefiles you downloaded. The bounding box above was pulled from the osm database, this bounding box is for the polygons you actually used. These two boxes may be the same or they may be different but the one derived from your downloaded shapefile is reproducible. 

Notice that we specify osm_multipolygons instead of osm_polygons. This is a case-specific choice. When you download a shapefile from OSM, if will include a number of different spatial object types and you can choose several of those options to move forward. We chose multipolygons because they best matched our area of interest. Use the quick plot below to visualize your area of interest before continuing. 
```{r pulled bounding box small, cache=TRUE}
boundaries <- my_boundary$osm_multipolygons[4,] #change to multipolygons
pulled_bb <- st_bbox(boundaries)
pulled_bb
```

This quick map allows you to check that the polygon you downloaded matches your expectations. 
```{r plot of area of interest small, cache=TRUE, warning=FALSE, fig.height=8}
basemap <- ggplot(data = boundaries) +
  geom_sf(fill = "cornflowerblue") +
  geom_sf_text(aes(label = boundaries$name)) 

basemap
```

# Download full time series from a single point
WARNING: TIME CONSUMING STEPS AHEAD

It's time to download. Here is a simple and fast example of data from a single point (the centroid of our polygon) for one year (2099). On our local machine this takes ~3 seconds. 

```{r pull data for single point for one year, eval=TRUE}
start_time <- Sys.time()

center_point <- st_centroid(boundaries) %>% st_bbox(center_point)
times <- inputs$available_times

Pulled_data_single_space_single_timepoint <- inputs$src %>% 
  hyper_filter(lat = lat <= c(center_point[4]+0.05) & lat >= c(center_point[2]-0.05)) %>% 
  hyper_filter(lon = lon <= c(center_point[3]+0.05) & lon >= c(center_point[1]-0.05)) %>%
  hyper_filter(time = times$`Available times` ==  44558) %>% 
  hyper_tibble(select_var = input_variables[1:38]) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, agr = "constant") 

end_time <- Sys.time()
print(end_time - start_time)

head(Pulled_data_single_space_single_timepoint)
```
Error note: an error like this is likely a connection error. Try again. Sometime you need to restart your R session to clear memory.
"curl error details: 
Error in R_nc4_open: NetCDF: I/O failure
Error in ncdf4::nc_open(x$source$source[1]) : 
  Error in nc_open trying to open file https://cida.usgs.gov/thredds/dodsC/macav2metdata_daily_future"
  
Here we remove the time filter to download all the requested input variables for the entire timeperiod.

NOTE: When you ask for a big download, that may exceed the API limit, the API asks you to confirm that you want to try. This will prompt a YES/NO question in your console that you need to answer if you want to continue. If you find this step taking a very long time, check that your console isn't waiting for you to confirm that you want to try this request. 

```{r pull data for single point for all years, eval=FALSE}
start_time <- Sys.time()
center_point <- st_centroid(boundaries) %>% st_bbox(center_point)

Pulled_data_single_space_all_timepoints <- inputs$src %>% 
  hyper_filter(lat = lat <= c(center_point[4]+0.05) & lat >= c(center_point[2]-0.05)) %>% 
  hyper_filter(lon = lon <= c(center_point[3]+0.05) & lon >= c(center_point[1]-0.05)) %>%
  hyper_filter(time = input_times$`Available times` ==  44558) %>% 
  hyper_tibble(select_var = input_variables) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, agr = "constant")

end_time <- Sys.time()
print(end_time - start_time)

head(Pulled_data_single_space_all_timepoints)
```

## Error when you try to download too much data at one time
  This is where you will start seeing error returned from the API if you have requested too much data at one time. 
  
  Downloads from the API are limited to 500mb per request. A request for a large area of interest combined with a long time series will return a cryptic runtime error informing you that your request was too large.

This error may look like this:
"CURL Error: Transferred a partial file
Error in Rsx_nc4_get_vara_int: NetCDF: DAP failure
Var: pr_CCSM4_r6i1p1_rcp85 Ndims: 3 Start: 0,444,511 Count: 34333,2,3
Error in ncvar_get_inner(ncid2use, varid2use, nc$var[[li]]$missval, addOffset, :
C function Rsx_nc4_get_var_int returned error"

The solution to this problem is to subset your requests to make them fit within the boundaries of the API. You can achieve this by balancing the size of your requested spatial extent and the length of your requested time period. For a small national park, it's possible to pull the entire time series from the API but larger parks will require you to request shorter time window or stitch multiple time windows together.  


## Filter time
Time is one of the easiest dimensions to filter along because the time data are organized by sequential number, so they are easy to cut up and reassemble without needing to reference outside tables. The request filter is sensitive to vector length so, we build in our filter indexes into the time dataframe to insure that they are all the same length. 

Here we are adding a generic index filled with 0's that we can use in the future, a column with 1's for the first half and 0's for the second half, and a column with 0's for the first half and 1's for the second. 
```{r filter time for single poing with error, cache=TRUE, eval=TRUE}
# Year 2034
time_min <- 38716
time_max <- 73048

input_times <- inputs$available_times %>% 
  add_column(index = 0) %>% 
  add_column(first_half = 0) %>% 
  add_column(second_half = 0) 
input_times[which(inputs$available_times[,1] >= time_min & inputs$available_times[,1] <= time_max ),3] <- 1

med <- median(row_number(input_times[,3])) 
input_times[which(as.numeric(row.names(input_times)) <= med),4] <- 1
input_times[which(as.numeric(row.names(input_times)) > med),5] <- 1

tail(input_times)
```

##Download full time series as a set of smaller downloads
When your data request is too big for a single API request, it will need to be cut it into smaller requests. You can use the new index columns we created in the time table above to cut our single request into two requests. This leaves you with two data tables instead of one. You can bind these two tables back together, but it is time consuming to do so because the two tables have to take turns writting themselves to a new third table that they can occupy together. This takes a lot of short term memory and errors out some machines. It will save you time and effort if you can avoid binding them back together. 
```{r stitch_pull for single poing, cache=TRUE, eval=FALSE}
start_time <- Sys.time()

Pulled_data_sub1 <- Pulled_data <- inputs$src %>% 
  hyper_filter(lat = lat <= c(center_point[4]+0.05) & lat >= c(center_point[2]-0.05)) %>% 
  hyper_filter(lon = lon <= c(center_point[3]+0.05) & lon >= c(center_point[1]-0.05)) %>% 
  hyper_filter(time =  input_times[,4] == 1) %>% 
  hyper_tibble(select_var = input_variables
    )

Pulled_data_sub2 <- Pulled_data <- inputs$src %>% 
  hyper_filter(lat = lat <= c(center_point[4]+0.05) & lat >= c(center_point[2]-0.05)) %>% 
  hyper_filter(lon = lon <= c(center_point[3]+0.05) & lon >= c(center_point[1]-0.05)) %>% 
  hyper_filter(time =  input_times[,5] == 1) %>% 
  hyper_tibble(select_var = input_variables
    )

end_time <- Sys.time()
end_time - start_time

tail(Pulled_data_sub1)
tail(Pulled_data_sub2)
```


# Expanding to include a large area of interest
In the previous examples, we pulled data from a single point in the center of a small national park. Now, we will show you how to download, analyze, and visualize these data over a large spatial extent.

First, Let's redefine our area of interest to include all the national parks that touch Yellowstone National Park. OSM provides a bounding box just like we used above for Hot Springs, Arkansas, but the national parks service also provides an official bounding box for this area of interest that we can enter manually if we like. This manually provided bounding box the specifies the area for extracting polygons from OSM but we still recalibarate that bounding box after download so it matches the downloaded polygons. 
```{r bounding box large, cache=TRUE}
bb <- getbb("yellowstone")

bb_manual <- bb
bb_manual[1,1] <- -111.15594815937659
bb_manual[1,2] <- -109.8305463801207
bb_manual[2,1] <- 44.12354048271325
bb_manual[2,2] <- 45.11911641599412

my_boundary_yellow <- opq(bb_manual) %>% 
  add_osm_feature(key = "boundary", value = "national_park") %>% 
osmdata_sf() 

my_boundary_yellow

boundaries_large <- my_boundary_yellow$osm_multipolygons
pulled_bb_large <- st_bbox(boundaries_large)
pulled_bb_large
```

Let's do another quickplot to make sure our area of interst matches our expectiaions. 
```{r plot of area of interest large, cache=TRUE, warning=FALSE, fig.height=8}
basemap <- ggplot(data = boundaries_large) +
  geom_sf(fill = "cornflowerblue") +
  geom_sf_text(aes(label = boundaries_large$name)) 

basemap
```

## Large area with few variables
  As you area of interest expands, you will need to cut you time and variable requests to accommodate more spatial points. Here we have cut our variables down to just "Precipitation", our scenarios down to just "RCP 8.5", and our models down to just "Model for Interdisciplinary Research On Climate 5"
```{r filter variables few, cache=TRUE}
input_variables <- inputs$variable_names %>% 
  filter(Variable %in% c( "Precipitation")) %>% 
  filter(Scenario %in% c( "RCP 8.5")) %>% 
  filter(Model %in% c(
             
 "Model for Interdisciplinary Research On Climate 5" )) %>%
  
  pull("Available variable")

input_variables
```

# Download data by AOI, filtered times, and filtered variable list
```{r pulled data for small area, cache=TRUE, eval=TRUE}
start_time <- Sys.time()

Pulled_data_large_area_few_variables <- inputs$src %>% 
  hyper_filter(lat = lat <= c(pulled_bb_large[4]+0.05) & lat >= c(pulled_bb_large[2]-0.05)) %>% 
  hyper_filter(lon = lon <= c(pulled_bb_large[3]+0.05) & lon >= c(pulled_bb_large[1]-0.05)) %>%
  hyper_filter(time = input_times$`Available times` ==  73048) %>% 
  hyper_tibble(select_var = input_variables
    ) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, agr = "constant")



end_time <- Sys.time()
end_time - start_time


head(Pulled_data_large_area_few_variables)
tail(Pulled_data_large_area_few_variables)
```
### Time extent plot
Check this plot to make sure you downloaded your entire time series. 
```{r plot pulled data for small area, eval=FALSE}
plot(Pulled_data$time, Pulled_data$`pr_MIROC5_r1i1p1_rcp85`)
```

### Spatial extent plot
Check here to make sure you downloaded the proper spatial extent. 
```{r check pulled data for large area, cache=TRUE, fig.height=8, eval=TRUE}
check_filter <- Pulled_data_large_area_few_variables %>% filter(time == min(Pulled_data_large_area_few_variables$time))

ggplot() +
  geom_sf(data = boundaries_large, fill = "cornflowerblue") +
 geom_sf(data = check_filter, color = "red", size=0.5) +
  coord_sf(crs = 4326) 
```




## If you encounter an error suggesting that you are pulling too much data, you will need to stitch a few requests together to keep each call below the 500mb limit. 

```{r stitch_pull for large area, cache=TRUE, eval=FALSE}
start_time <- Sys.time()


Pulled_data_sub1 <- inputs$src %>% 
  hyper_filter(lat = lat <= c(pulled_bb[4]+0.05) & lat >= c(pulled_bb[2]-0.05)) %>% 
  hyper_filter(lon = lon <= c(pulled_bb[3]+0.05) & lon >= c(pulled_bb[1]-0.05)) %>% 
  hyper_filter(time =  input_times[,4] == 1) %>% 
  hyper_tibble(select_var = input_variables) %>%  
  st_as_sf(coords = c("lon", "lat"), crs = 4326, agr = "constant")
## should time be in here?


Pulled_data_sub2 <- inputs$src %>% 
  hyper_filter(lat = lat <= c(pulled_bb[4]+0.05) & lat >= c(pulled_bb[2]-0.05)) %>% 
  hyper_filter(lon = lon <= c(pulled_bb[3]+0.05) & lon >= c(pulled_bb[1]-0.05)) %>% 
  hyper_filter(time =  input_times[,5] == 1) %>% 
  hyper_tibble(select_var = input_variables) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326, agr = "constant")
## should time be in here?

end_time <- Sys.time()
end_time - start_time



tail(Pulled_data_sub1)
tail(Pulled_data_sub2)
```

```{r rbind pulled results from large area, cache=TRUE, eval=FALSE}
start_time <- Sys.time()

Pulled_data_stitch <- bind_rows(Pulled_data_sub1, Pulled_data_sub2)

end_time <- Sys.time()
end_time - start_time
```

### Time extent plot
Check this plot to make sure you downloaded your entire time series. 
```{r plot stitched data, cache=TRUE, eval=FALSE}
plot(Pulled_data_stitch$time, Pulled_data_stitch$`pr_MIROC5_r1i1p1_rcp85`)
```


# Melt downloaded points into a raster before aggretation
```{r rasterize with stars, cache=TRUE, fig.height=7}
rast <- st_rasterize(Pulled_data_large_area_few_variables) 
plot(rast)

#Pulled_data %>% as.data.frame() %>% brick()
```


```{r GridMet view, include=FALSE, eval=FALSE}
param_meta$gridmet
```


```{r GridMet, eval=FALSE,fig.height=8, include=FALSE}
subed_times <- input_times %>% filter(index == 1) 
GM <- getGridMET(st_as_sfc(st_bbox(boundaries)), "tmax", startDate = "1997-04-06", endDate = "1999-12-30")
SM_stars <- GM$gridmet_tmax %>% brick() %>%  st_as_stars()
#st_set_dimensions(SM_stars, 3, values = X1997.04.06, names = "tmax")

ggplot() +
  geom_sf(data = SM_stars, fill = "cornflowerblue") +
 geom_sf(data = check_filter, color = "red", size=0.5) +
  coord_sf(crs = 4326) 
```

```{r, eval=FALSE, include=FALSE}
st_get_dimension_values(rast)
#combo <- st_extract(rast, SM_stars) %>% st_as_sf()
combo <- c(SM_stars,rast , along_crs=TRUE, along=c(1,2))
class(combo) 
class(SM_stars)
plot(combo$X2000.01.01)
```

```{r, eval=FALSE, include=FALSE}
plot(combo$X2000.01.01)

#extracted_GridMET <- st_extract(rast, SM_stars) %>% st_as_sf()
ggplot(data=combo) +
  geom_sf(aes(fill = X2000.01.01)) +
  scale_fill_continuous(low="thistle2", high="darkred", 
                       guide="colorbar",na.value="white")+
  coord_sf(crs = 4326)

```



#Aggregate downloaded data to different spatial objects


## Aggregate to polygon (faster method)
```{r aggregate to polygon, cache=TRUE}
extracted <- st_extract(rast, boundaries_large$geometry) %>% st_as_sf()
names(extracted)[1] <- "nn"
ggplot(data=extracted) +
  geom_sf(aes(fill = nn)) +
  scale_fill_continuous(low="thistle2", high="darkred", 
                       guide="colorbar",na.value="white")+
  coord_sf(crs = 4326)
```


## Clip raster with polygon (slower method)
```{r cut out raster with polygon, cache=TRUE}
intersection <- st_intersection(x = Pulled_data_large_area_few_variables, y = boundaries_large$geometry)

names(intersection)[1:2] <- c("Precipitation","b")
```


```{r plot of raster mask, cache=TRUE}
library(ggthemes)
ggplot() +
  geom_sf(data = intersection, aes(color=Precipitation)) +
  scale_color_continuous(low="thistle2", high="darkred", 
                       guide="colorbar",na.value="white")+
  geom_sf(data = boundaries_large, fill = NA, color = "white") +
  theme_tufte()+
  labs(title = "YELLOWSTONE NATIONAL PARK", subtitle = "Temperture in 2050")
```


## Aggregate to River segment
```{r pull river data, fig.height=10, cache=TRUE}
river <- opq(bb_manual) %>%
  add_osm_feature(key = "waterway", value = "river") %>%
  osmdata_sf() 
river
```

```{r aggregate data to river lines, cache=TRUE}
river_sub <- st_buffer(river$osm_lines, 2200)
extracted_river <- st_extract(rast,  river_sub$geometry ) %>% st_as_sf()
head(extracted_river)
#colnames(extracted_river)[1] <- "var1"
colnames(extracted_river)[1] <- "pre"

```

```{r plot river aggregation, fig.height=8, cache=TRUE}
ggplot(data=extracted_river) +
  geom_sf(aes(fill = pre), size=0) +
   coord_sf(crs = 4326, xlim = c(pulled_bb_large[1], pulled_bb_large[3]), 
           ylim = c(pulled_bb_large[2], pulled_bb_large[4]),
           expand = FALSE) +
  scale_fill_continuous(low="thistle2", high="darkred", 
                       guide="colorbar",na.value="white")+
  labs(title = "Rivers of Yellowstone",
       subtitle = "Projected humidity in 2040", 
       caption = "Data Source: Climate Futures...") + 
  theme_tufte()
```

## Aggregate to road segment
```{r pull road data, fig.height=8, cache=TRUE}
roads <- opq(pulled_bb_large) %>%
  add_osm_feature(key = 'highway', value = 'primary') %>%
  add_osm_feature(key = 'highway', value = 'secondary') %>%
  osmdata_sf() 
roads_sub <- st_buffer(roads$osm_lines, 2200)
extracted_roads <- st_extract(rast,  roads_sub$geometry ) %>% st_as_sf()
extracted_roads

colnames(extracted_roads)[1] <- "pre"
```


```{r plot road aggregation, fig.height=8, cache=TRUE}
ggplot(data=extracted_roads) +
  geom_sf(aes(fill = pre), size=0) +
   coord_sf(crs = 4326) +
  scale_fill_continuous(low="thistle2", high="darkred", 
                       guide="colorbar",na.value="white")+
  labs(title = "Roads of Yellowstone",
       subtitle = "Projected humidity in 2040", 
       caption = "Data Source: Climate Futures...") + 
  theme_tufte()
```

### Computing new daily climate variables

Now that we have all of the climate parameters for our study region, we can
compute functions of those variables. 
For example, it is common to compute the midpoint of the maximum and minimum 
daily temperature, which we can do using the `mutate` function: 

```{r temp-midpoint, eval=FALSE}
df <- df %>%
  mutate(tasmid = (tasmax + tasmin) / 2)
```

Now we have a new column called `tasmid` that is the midpoint of the maximum
and minumum daily temperature!

Wind speed provides another example of a derived parameter that can be computed 
for each day. 
By default, we have two wind-related parameters: the eastward wind component
(called `uas`) and the northward wind component (called `vas`), both in units of 
meters per second (you can get this information from `cft::argument_reference`). 
Wind speed can be computed from `vas` and `uas` using the Pythagorean theorem: 

$\text{Wind speed} = \sqrt{v_{as}^2 + u_{as}^2}.$

In code: 

```{r wind-speed, eval=FALSE}
df <- df %>%
  mutate(wind_speed = sqrt(vas^2 + uas^2))
```


### Computing new climate variable summaries

Sometimes, there are new climate variables that summarize daily data. 
For example, you may want to compute: 

- Last Day of Frost (i.e., last day in spring when min. air temp. < 0 C)
- First Day of Frost (i.e., first day in fall when min. air temp. < 0 C)
- Number of days above or below some threshold (e.g., days with max. air temperature over 40 C, or days with > 1mm of precipitation)
- Growing season length (# days with air temperature > 0 C)

All of these quantities summarize daily data, and require some aggregation time interval which in many cases will be one year. 
As an example, we will compute the growing season length for Wind Cave National Park across all models and emissions scenarios. 
To do this, we first need to define a new column for year, which we will use as a grouping variable: 

```{r get-year, eval=FALSE}
df <- df %>%
  mutate(year = year(date))
```

Now, we want to compute growing season length for each year, model, emissions scenario combination. 

```{r grow-season, eval=FALSE}
growing_seasons <- df %>%
  group_by(rcp, model, year, ensemble) %>%
  summarize(season_length = sum(tasmid > 273.15)) %>%
  ungroup
```

Notice that we used our derived temperature midpoint column `tasmid`, and computed the total (`sum()`) number of days for each group where the temperature midpoint was greater than 0 C (or, 273.15 Kelvin, which are the units of the temperature data).

```{r glimpse-grow-season, eval=FALSE}
growing_seasons
```

Let's visualize the growing season over time for each model and emission scenario:

```{r plot-grow-season, fig.height = 5, fig.width = 6, eval=FALSE}
growing_seasons %>%
  ggplot(aes(year, season_length, color = rcp, group = model)) + 
  geom_line(alpha = .3) + 
  facet_wrap(~rcp, ncol = 1) + 
  xlab("Year") + 
  ylab("Growing season length (days)") + 
  scale_color_manual(values = c("dodgerblue", "red")) + 
  theme(legend.position = "none")
```


## Comparing climate in two time periods

Use the tibble object that is returned from `cft_df()` as an input to
`compare_periods()` to compare climate between a reference and target period. You
may specify the function with which to aggregate your chosen variable as well
as the yearly time period months of the year to include in this calculation.

```{r comps, eval=FALSE}
comps <- compare_periods(df,
                         var1 = "pr",
                         var2 = "tasmax",
                         agg_fun = "mean",
                         target_period = c(2025, 2030),
                         reference_period = c(2020, 2024),
                         months1 = 5:8,
                         months2 = 5:8,
                         scenarios = c("rcp45", "rcp85"))
```

This provides a data frame that can be used to compare the values in the target
and reference period.

```{r glimpse-comps, eval=FALSE}
glimpse(comps)
```

One useful plot shows the difference in the two variables between reference and
target periods:

```{r plot-comps, fig.height = 6, fig.width = 9, eval=FALSE}
title <-  paste("Change from the historical vs. reference period:", 
                comps$reference_period, comps$target_period, sep= "  vs  " )[1]

comps %>%
  dplyr::select(parameter, rcp, model, reference_period, target_period, difference) %>%
  pivot_wider(names_from = parameter, values_from = difference) %>%
  ungroup %>%
  mutate(rcp = ifelse(rcp == "rcp45", "RCP 4.5", "RCP 8.5")) %>%
  ggplot(aes(pr, tasmax, color = rcp)) + 
  ggtitle(title) +
  geom_point() + 
  geom_hline(yintercept = 0, alpha = .2) + 
  geom_vline(xintercept = 0, alpha = .2) +
  geom_text_repel(aes(label = model), segment.size = .3, size = 3) + 
  xlab("Difference in mean daily precipitation (mm)") + 
  ylab("Difference in mean daily max. temperature (C)") + 
  scale_color_manual(values = c("dodgerblue", "red"), 
                     "Greenhouse gas\ntrajectory") 
```

So, nearly all model runs indicate warming, but the amount of warming varies by 
model and emissions scenario. 
Precipitation increases and decreases are predicted by different models. 




