---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%", 
  dpi=300
)
```

# cst: climate scenarios toolkit

<!-- badges: start -->
[![Build Status](https://travis-ci.com/earthlab/cst.svg?branch=master)](https://travis-ci.com/earthlab/cst)
[![codecov](https://codecov.io/gh/earthlab/cst/branch/master/graph/badge.svg)](https://codecov.io/gh/earthlab/cst)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
<!-- badges: end -->

The goal of cst is to provide data access utilities that facilitate climate 
scenario planning

## Installation

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("earthlab/cst")
```
## Example

To acquire precipitation data for Acadia National Park for a subset of climate
models, you can use the `cstdata` function:

```{r}
library(ggplot2)
library(cst)

d <- cstdata(park = "Acadia National Park", parameters = "pr", 
             years = c(2020, 2021), models = "CCSM4", scenarios = "rcp85")
```

This gives you a data frame with paths to local climate data files: 

```{r print-d}
str(d)
```

And, you can also summarize the daily data by computing a spatial average over the region of interest: 

```{r summarize-data}
df <- cst_df(d, cores = 2)
str(df)
```

Because this is a data.frame, you can use all of the normal data visualization and processing functionality in base R!

```{r base-plot}
plot(x = df$date, 
     y = df$pr, 
     type = "b", 
     ylab = "Precipitation (mm)", 
     xlab = "Date", 
     main = "Acadia National Park, CCSM4, RCP 8.5")
```

Or, if you prefer, you can use the tidyverse: 

```{r ggplot-precip}
df %>%
  ggplot(aes(date, pr)) + 
  geom_point() + 
  geom_line() + 
  xlab("Date") + 
  ylab("Precipitation (mm)") + 
  ggtitle("Acadia National Park, CCSM4, RCP 8.5")
```

This is just a small glimpse at what you can do with the cst package.
For more, see [Getting started with the Climate Scenarios Toolkit](vignettes/cst-intro.Rmd)

## Meta

* Please [report any issues or bugs](https://github.com/earthlab/cst/issues),
after reading our contribution [guidelines](CONTRIBUTING.md), and the 
[Contributor Code of Conduct](CONDUCT.md). 
* License: GPL-3
* See `citation("cst")` in R to cite this package in publications. 