---
title: "cstdata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cstdata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Climate Scenarios Toolbox

The use of Global Circulation Models (GCMs) allows resource managers and
researchers to study/plan for possible future climate conditions. The
complexity of the climate system has lead to a large number such GCMs and it is
common practice to examine outputs from many different models at once. Using
methods such as the ensemble mean of these outputs helps to account for
uncertainties inherent in any single model while the outputs of single models
can give an idea of the range of possible future conditions.

The spatial scale of the original outputs of these models is usually quite
coarse (often 1 degree), so several downscaling methods have been developed
to allow for smaller-scale study. One such method is called the Multivariate
Adaptive Climate Analog (MACA).{desription here}(link)[here].

Because the amount of data generated by downscaled GCMs can be quite large
(at least daily time-scales and 2-4 km spatial resolutions), the Climate
Scenarios Toolbox was developed to help users quickly download smaller subsets.
It will also use these subsets in a variety of plotting functions.

Data is acquired from the Northwest Knowledge Server of the University of
Idaho (link)[here]. While this package only accomdates  {what are the
limits of availability}, this accounts for ~20 TB of data access. 


## cstdata

The cstdata packages accomplishes three goals; acquiring, subsetting, and
plotting outputs from MACA-downscaled Global Circulation Models (GCMs) outputs
from the Northwest Knowledge Network (link here). 

```{r library}
library(cstdata)
```

* Acquiring and subsetting data

Use the cstdata reference object within the cstdata package to learn which
GCMs are available. 

```{r references}
references <- cstdata::argument_reference
print(references$models)
```

This package was originally writted with the National Park Service in mind, so it Combine this with a location in the form of a shapefile path or a US National Park Name to call the cstdata
function and download the data. You may specify a range of years, a set of models, a set of parameters, and a
set of representative concentration pathways to return. Leaving these arguments empty will results in a download
of all available data for that location.

```{r download_1}
file_refs <- cstdata(park = "Wind Cave National Park",
                     years = c(1980, 2050),
                     models = NA,
                     parameters = c("tasmin", "vpd"),
                     scenarios = c("rcp45", "rcp85"),
                     ncores = 4)
```


* Using a shapefile to subset data.

You may also specify a shapefile as a local path or as a remote path to a zipped file.

``` {r download_2}

```

* Storing outputs in an s3 bucket

You may choose to save results to an Amazon Web Services s3 bucket. To do so,
you will need to set the following information about your s3 bucket as
environment variables:

      "BUCKET_NAME" ?
      "AWS_ACCESS_KEY_ID"
      "AWS_SECRET_ACCESS_KEY"
      "AWS_DEFAULT_REGION"

This can be done at the beginning of a session with the following command: ?

`sys.setenv(c("BUCKET_NAME" = <bucket name>, ?
              "AWS_ACCESS_KEY_ID" = <value>,
              "AWS_SECRET_ACCESS_KEY" = <value>,
              "AWS_DEFAULT_REGION" = <value>
              )
            )`

To automate this ...
If you have forgotten your access key or it has expired...
If you have forgotten you secret access key, you will have to reset it here...

After setting your credentials, set the `s3_bucket` argument to true and run
the function normally.

```{r download_3}
# proj_dir = "."
# file_refs <- cstdata(park = "Wind Cave National Park",
#                      years = c(1980, 2050),
#                      models = NA,
#                      parameters = c("tasmin", "vpd"),
#                      scenarios = c("rcp45", "rcp85"),
#                      local_dir = proj_dir,
#                      s3_bucket = TRUE)
```

## scatterplot

Use the file reference object that is returned from cstdata as an input to
`scatterplot` (we need to change the name I think, there's another one) to
create a scatterplot of two variables of your choice from multiple GCMs. You
may specify the function with which to aggregate your chosen variable as well
as the yearly time period months of the year to include in this calculation.
If you would like to see the difference in values between a two periods, set
the `difference` argument to TRUE and provide a reference period with the
`reference_period` argument

```{r scatterplot}
graph_data <- scatterplot(file_refs,
                          var1 = "tasmin",
                          var2 = "vpd",
                          agg_fun = "mean",
                          difference = TRUE,
                          target_period = c(2020, 2050),
                          reference_period = c(1990, 2019),
                          months1 = c(1, 2, 3),
                          months2 = c(4, 5, 6),
                          scenarios = c("rcp45", "rcp85"))
```

## boxplot
``` {r boxplot}
```

# timeseries
``` {r timeseries}

```

# raw_data
``` {r raw data}
```